% ASCII-only LaTeX. Evidence cutoff 2026-07-15.
\documentclass[11pt]{article}
\usepackage[margin=0.85in]{geometry}
\usepackage{booktabs,longtable,array,amsmath,amssymb,hyperref,microtype}
\hypersetup{hidelinks}
\emergencystretch=3em
\newcommand{\observed}{\textbf{Observed:}}
\newcommand{\inferred}{\textbf{Inferred:}}
\newcommand{\limitn}{\textbf{Limitation:}}
\title{Chaos Language Algorithm Experiments\\A Comprehensive Evidence-Bounded Report}
\author{Chaos Language Algorithm / OmegaSim projects}
\date{Evidence cutoff: July 15, 2026}

\begin{document}
\maketitle

\begin{abstract}
This report consolidates every durable Chaos Language Algorithm (CLA)
experiment found in the project and repository records through July 15, 2026,
with emphasis on the newer high-dimensional work absent from earlier reports.
CLA currently provides exact-reconstructing symbolic grammar induction,
low-cardinality and fixed symbolization, an approximate two-part compression
proxy, shuffled-surrogate comparisons, and held-out n-gram diagnostics. Across
the evidence, exact reconstruction is reliable, raw rectangular partitions
fail in high dimension, and low-cardinality microstates restore recurrence.
However, positive compression margins are not unique to VAMP embeddings, and a
leakage-free rank-conditioned comparison rejected further linear-VAMP tuning:
VAMP beat matched PCA controls on both primary metrics in zero of three
trajectories at both alphabet sizes. The report therefore separates positive
engineering results, detector failures, null comparisons, and hypotheses; it
does not claim a validated chaos classifier or calibrated CLA likelihood.
\end{abstract}

\section{Scope, sources, and epistemic discipline}

The canonical code is the \texttt{chaoslang} repository on branch
\path{agent/hd-embedding-cla}; the report build uses commit \texttt{974af31}
plus this document. OmegaSim detector integration uses branch
\path{agent/cla-detector-prereg}, commit \texttt{5a6002b}. Durable run records
are under \path{projects/chaos-language-algorithm/experiments}, repository
\path{experiments}, and \path{projects/omegasim/experiments}.

\observed{} denotes direct code, artifact, or recorded-run evidence.
\inferred{} denotes interpretation that follows from those observations.
Proxy quantities retain historical ``bits'' field names where necessary, but
the current scorer is not a complete calibrated code. Held-out perplexity is
from a smoothed n-gram model over symbols, not from a predictive CLA grammar.

\section{Detector pipeline and what each metric establishes}

The experimental pipeline is
\[
  x_{1:n} \rightarrow \hbox{symbolizer} \rightarrow s_{1:n}
  \rightarrow \hbox{CLA grammar and exact parse}.
\]
CLA greedily proposes repeated chunks and contextual categories while requiring
the final grammar/parse to expand exactly to the original symbol stream.

\begin{longtable}{p{0.23\linewidth}p{0.34\linewidth}p{0.35\linewidth}}
\toprule
Measurement & What it directly checks & What it does not establish \\
\midrule
\endhead
Exact reconstruction & Grammar and parse losslessly reproduce the observed
symbols & Generating partition, chaos, prediction, or useful abstraction \\
Compression proxy & Current grammar representation is shorter/longer than an
empirical-symbol baseline under the implemented proxy & A calibrated MDL code,
likelihood, or significance level \\
Real-minus-surrogate margin & Temporal order helps relative to a specified
shuffle/block surrogate & Uniqueness to the proposed embedding or dynamics \\
Held-out n-gram loss & Symbol stream has train-to-test finite-order predictive
regularity & Predictive performance of the CLA grammar itself \\
Embedding singular values & Numerical modes returned by a kinetic estimator &
Validated slow processes when modes are near one or conditioning is weak \\
\bottomrule
\end{longtable}

\section{Baseline attractor and scaling experiments}

The first lane used an equal-width rectangular partition (M1) and a small greedy
CLA budget. All recorded runs reconstructed exactly. Table~\ref{tab:m1}
consolidates the baseline records.

\begin{table}[ht]
\centering\small
\begin{tabular}{lrrrrr}
\toprule
System & Dim. & Samples & Unique symbols & Rules & Proxy score \\
\midrule
Mackey--Glass & 1 & 256 & 4 & 6 & 44.4 \\
Logistic map & 1 & 256 & 4 & 8 & 113.2 \\
Lorenz--63 & 3 & 256 & 21 & 8 & 187.2 \\
Rossler & 3 & 256 & 24 & 8 & 189.2 \\
Lorenz--96 & 5 & 256 & 56 & 8 & 234.2 \\
Lorenz--96 & 8 & 256 & 78 & 8 & 229.2 \\
Lorenz--96 & 12 & 256 & 107 & 6 & 241.4 \\
Lorenz--96 & 16 & 256 & 121 & 3 & 248.7 \\
Lorenz--96 & 20 & 256 & 127 & 6 & 251.4 \\
Lorenz--96 & 24 & 256 & 148 & 5 & 251.5 \\
Lorenz--96 trie/JS & 20 & 1024 & 510 & 8 & 1021.2 \\
\bottomrule
\end{tabular}
\caption{Recorded M1 experiments. Scores are proxy units, not calibrated bits.}
\label{tab:m1}
\end{table}

\observed{} In the strongest high-dimensional M1 record, 1024 samples of
20-dimensional Lorenz--96 yielded 510 unique compound symbols, eight accepted
chunk rules, no category, score 1021.2, and 2.55 seconds fit time. Provenance:
\texttt{20260709T192728Z-lorenz96-1024-dim20-suffix-trie} at commit
\texttt{b344aa3c}. The score stayed close to raw sequence length.

\inferred{} The main failure occurred before grammar induction: the Cartesian
partition produced an alphabet too sparse for robust recurrence. The run is not
evidence that Lorenz--96 lacks symbolic grammar. It motivated choosing alphabet
cardinality directly with microstates.

\section{Lifted Lorenz--63 Phase-1 experiment}

The first adaptive high-dimensional experiment lifted Lorenz--63 into
$\mathbb{R}^{256}$ with small deterministic noise, used a dependency-free
kinetic map to dimension three, clustered microstates, and compared with raw M1.

\observed{} In the one-seed diagnostic, adaptive microstates achieved a
real-minus-shuffled compression proxy of 39.33 versus 2.10 for M1. Held-out
n-gram perplexity was 3.42 versus 433.18. Exact reconstruction held. Provenance:
\texttt{20260712T200000Z-lorenz63-lift256-phase1}.

\inferred{} Low-cardinality symbolization decisively repaired the M1 alphabet
explosion on this fixture. The run did not isolate the kinetic embedding from
clustering, used one trajectory, and did not compare against direct raw-space
k-means. It therefore supported the pipeline seam, not VAMP superiority.

\section{Lifted Lorenz--63 multi-seed VAMP sweep}

The next preregistered sweep used three seeds, a 192-point retained trajectory,
a noisy $\mathbb{R}^{256}$ lift, deeptime VAMP dimensions $\{3,5\}$, alphabet
sizes $\{8,16,24\}$, and lags $\{1,3\}$. It produced 36 VAMP rows and nine
controls.

\begin{table}[ht]
\centering\small
\begin{tabular}{lrrr}
\toprule
Method/aggregate & Rows & Mean proxy margin & Mean n-gram perplexity \\
\midrule
deeptime VAMP overall & 36 & 8.085 & 11.628 \\
direct raw xyz k-means & 3 & 12.200 & 3.697 \\
dependency-free reference & 3 & 15.167 & 4.151 \\
raw compound M1 & 3 & 1.867 & 257.769 \\
\bottomrule
\end{tabular}
\caption{Aggregate multi-seed results; proxy margins are not calibrated MDL.}
\end{table}

\observed{} All 36 VAMP configurations had positive real-minus-shuffled proxy
margins, but direct xyz k-means and the dependency-free reference were as good
or better on the aggregate metrics. No categories emerged. The recorded
near-unit singular values exposed a diagnostics bug: finite-only validation had
failed to flag invalid or implausible implied timescales. The writer was repaired
to preserve raw values and fail closed. Provenance:
\texttt{20260712T200200Z-lorenz63-r256-deeptime-multiseed}, commits
\texttt{bee5965} and \texttt{64e9dd6}.

\inferred{} Positive shuffled-surrogate margins were a property of many
low-cardinality streams, not evidence uniquely favoring VAMP. The correct next
comparison required leakage-free train-only transforms and alphabet-matched
PCA/raw controls.

\section{Leakage-free rank-conditioned VAMP/TICA test}

The decisive follow-up used three independent 2048-point Lorenz--63
trajectories, train-only noisy $\mathbb{R}^{256}$ lifts, 60/20/20 blocked splits
with 16-point lag gaps, PCA ranks $\{3,10\}$, VAMP/TICA dimensions $\{2,3\}$,
lags $\{1,4,16\}$, $k=\{8,16\}$, and 20 circular block surrogates at block
lengths 4 and 16. Selection occurred on blocked validation VAMP-2 and test was
evaluated once.

\begin{table}[ht]
\centering\small
\begin{tabular}{rrrrrr}
\toprule
Seed & $k$ & VAMP loss & best PCA loss & VAMP margin & best PCA margin \\
\midrule
101 & 8 & 0.434 & 0.420 & 5.20 & 3.65 \\
202 & 8 & 0.482 & 0.423 & 5.20 & 5.55 \\
303 & 8 & 0.615 & 0.543 & 13.00 & 7.30 \\
101 & 16 & 0.731 & 0.710 & 4.30 & 2.60 \\
202 & 16 & 0.771 & 0.727 & 4.25 & 3.05 \\
303 & 16 & 1.050 & 0.952 & 3.90 & 9.00 \\
\bottomrule
\end{tabular}
\caption{Test loss in bits/symbol; margin is the minimum proxy margin across
block lengths. Lower loss and higher margin are better.}
\end{table}

\observed{} VAMP beat the best matched PCA control on both metrics in zero of
three trajectories for $k=8$ and zero of three for $k=16$, versus the
preregistered requirement of at least two of three. VAMP test loss was worse in
every seed/alphabet stratum. All 72 estimator candidates were numerically valid;
selected modes were near-unit, so they were not treated as standalone kinetic
evidence. Runtime was 163 seconds and the full suite passed 115 tests.
Provenance: \texttt{20260713T153839Z-lorenz63-rank-conditioned-vamp-tica},
commit \texttt{4423f8f}.

\inferred{} Linear VAMP development correctly stopped. This bounded null result
does not reject nonlinear embeddings, other dynamics, or CLA; it rejects the
claim that the tested linear VAMP path improved on matched PCA for this lift.

\section{OmegaSim restart under a fresh detector preregistration}

Ben approved restarting OmegaSim on July 15 under a new preregistration. The
frozen design evaluates A6 gain 5.0, coupling $\{0.35,0.60\}$, delay $\{0,3\}$,
controls \{appraisal, linear, shuffled\}, five seeds, and separate 4D, 8D, and
20D feature strata. A deterministic $k=8$ microstate model is fit on the first
60 percent of each trajectory and frozen before evaluation. Each row uses five
marginal-preserving symbol surrogates.

The row-level detector requires exact reconstruction, at least four occupied
states, positive proxy margin, and held-out order-2 loss at least 0.10
bits/symbol below the surrogate mean. Promotion additionally requires appraisal
to beat both matched dynamics controls on both metrics in at least four of five
seeds, without pooling dimensionality strata. Full protocol:
\texttt{omegasim/docs/cla\_detector\_preregistration\_20260715.md}.

\observed{} The local CPU run completed all 180 rows in 31 minutes 26 seconds.
All row-level software invariants passed. Exactly one of the twelve frozen
tuple/stratum promotion cells met the four-of-five matched-seed criterion:
coupling 0.60, delay 3, and the 8D role stratum. Appraisal beat both linear and
shuffled controls on compression margin and held-out loss in four seeds; seed
29 missed because its compression margin (118.6 proxy bits) was below the
linear control (132.2), although its held-out loss was lower. No other cell
reached four wins; the next highest cells reached three.

\inferred{} This is bounded positive evidence that the frozen CLA feature path
distinguishes appraisal dynamics from the two exact-tuple controls in one
predeclared regime. It is not evidence that the dynamics form a strange
attractor, that roles are semantic grammar categories, or that CLA is a
calibrated chaos detector. The singled-out cell now requires an untouched
replication and the still-required Mackey--Glass/Lorenz--96 calibration battery.
Immutable artifacts have SHA-256
\texttt{2539b8381b8e85317a7ca4ae53643483153c8f3535dabbef282abec880902eb4}
(JSON) and
\texttt{5dd819af1b76b40811712b8452adab8c5bcb57f1066847bf56cb779a7bd6e729}
(CSV).

\section{Implemented but not yet decisive experimental lanes}

\begin{itemize}
\item Train-fitted deterministic k-means symbolization is now a first-class
leakage-safe object (commit \texttt{974af31}); functional tests establish no
test-data refit, not scientific superiority.
\item Frozen-suffix coding and deterministic model encodings were implemented
on a separate branch (commits \texttt{5be45e5}, \texttt{ce0e476}) with 140 full
tests passing. The preregistered held-out benchmark has not yet produced a
durable result, so calibrated predictive coding remains open.
\item PCCA+ soft memberships exist in the optional backend, but automatic
category seeding and a decisive category recovery experiment remain pending.
\item The bounded suffix trie has functional/node-bound tests and one 2.55 s
Lorenz--96 run, but no matched wall-time/peak-memory comparison against the old
brute-force miner.
\end{itemize}

\section{Current implementation validation}

\observed{} On July 15, the command
\path{PYTHONPATH=src python3 -m unittest discover -s tests -v} completed 126
tests in 47.394 seconds with status OK on the report branch. This includes the
optional deeptime-backed tests available in the local environment and the new
train-fitted k-means no-refit regressions. A passing software suite establishes
the declared invariants and fixture behavior; it does not convert the null
scientific comparisons above into positive evidence.

\section{Conclusions and claim ledger}

\begin{longtable}{p{0.42\linewidth}p{0.16\linewidth}p{0.34\linewidth}}
\toprule
Claim & Status & Evidence / qualification \\
\midrule
\endhead
CLA exactly reconstructs tested symbolic streams & Supported & All durable
baseline and high-dimensional runs plus regression tests \\
Raw rectangular M1 scales poorly in high dimension & Supported for tested
fixtures & 510 unique symbols in 1024 by 20D; near-raw proxy score \\
Low-cardinality microstates restore recurrence & Supported & Lifted one-seed
and multi-seed runs; direct controls also strong \\
Positive proxy margin uniquely indicates grammar-bearing embedding & Rejected &
Direct raw/reference controls were equally or more positive \\
Linear VAMP improves matched PCA on lifted Lorenz--63 & Rejected for frozen
protocol & Zero of three joint wins at both $k$ values \\
CLA is a validated chaos detector & Not established & No calibrated operating
curve, likelihood, or matched broad benchmark battery \\
CLA categories recover meaningful attractor lobes & Not established & No
categories in the main high-dimensional sweeps \\
OmegaSim appraisal dynamics differ under the frozen CLA proxy in one regime &
Bounded support & One of twelve cells promoted: coupling 0.60, delay 3, roles8,
with matched wins in four of five seeds; untouched replication required \\
\bottomrule
\end{longtable}

The central scientific lesson is methodological: symbolization and controls
dominate interpretation. Exact reconstruction is necessary but easy; positive
compression relative to a weak shuffle is insufficient; and a useful embedding
must beat alphabet-matched, train-only alternatives on held-out prediction and
temporal-grammar diagnostics simultaneously. The current evidence supports CLA
as an inspectable experimental grammar toolkit. It does not yet support CLA as
a calibrated strange-attractor classifier.

\section*{Reproduction pointers}

Each project-level experiment directory contains \texttt{RUN.md},
\texttt{command.sh}, environment/git captures, status, logs, raw JSON/CSV, and
checksums where available. The decisive records are:
\begin{itemize}
\item \texttt{20260709T192728Z-lorenz96-1024-dim20-suffix-trie}
\item \texttt{20260712T200000Z-lorenz63-lift256-phase1}
\item \texttt{20260712T200200Z-lorenz63-r256-deeptime-multiseed}
\item \texttt{20260713T153839Z-lorenz63-rank-conditioned-vamp-tica}
\item \texttt{20260715T153049Z-cla-detector-preregistered} (OmegaSim)
\end{itemize}

\end{document}
